% !TEX spellcheck = en_US
% !TEX spellcheck = LaTeX
\documentclass[letterpaper,english,10pt]{article}
\input{../../header}

\title{Lecture-11: Correlated Variables}


\begin{document}
\maketitle
\section{Asymptotic equipartition}
We can count the number of configurations $X  \in \sX^N$ with either a given type $q(x)$ or a given empirical average  $\bar{f}$ of some observable $f: \sX^N \to \R$. 
We say that `there are approximately $2^{NH(q)}$ sequences with type $q$'. 
\begin{prop}
 The number $\cN_{K,N}$ of sequences $X \in \sX^N$ which have a type belonging to the compact set $K \subset \sM(X)$ behaves as $\cN_{K,N} \ee 2^{NH(q^\ast)}$, where $q^\ast =\arg\max\{H(q): q \in  K\}$.
\end{prop}
\begin{proof} 
Let $p(x)$ be the uniform probability distribution on $\sX$, then $P_p\set{q \in K} = \frac{\cN_{K,N}}{\abs{\sX}^N}$. 
From Sanov's theorem, we know that $P_p\set{q \in K}  \ee e^{-N D(q^{\ast} \Vert p)}$. 
Since $p$ is uniform over $\sX$, we have $D(q \Vert p) = \E_q\log_2 q(x) + \E_q\log_2\abs{\sX} = -H(q) + \log_2\abs{\sX}$. 
Combining the above results, we get that 
\EQ{
\cN_{K,N} = \P_p(q \in K)2^{N \log_2\abs{\sX}} \ee 2^{NH(q^\ast)}.
} 
\end{proof}
As a consequence of Sanov's theorem, 
we know that for any \emph{i.i.d.} random sequence $X \in \sX^N$ of $N$  with the common probability distribution $p(x)$, the most probable type is $p(x)$ itself, and that deviations are exponentially rare in $N$. 
We expect that almost all the probability will be concentrated into sequences that have a type close to $p(x)$ in some sense. 
On the other hand, because of the above proposition, the number of such sequences is exponentially smaller than the total number of possible sequences $\abs{\sX}^N$. 

Let's define what is meant by a sequence having a type `close to $p(x)$'. 
Given a sequence $X$, we introduce the quantity empirical entropy defined as
\EQ{
r(X) \triangleq -\frac{1}{N}\log_2P_N(X) = -\frac{1}{N}\sum_{i=1}^N\log_2p(X_i).
}
Clearly, $\expect{r(X)} = H(p)$. 
\begin{defn} 
A random sequence $X \in \sX^N$ is called $\epsilon$\textbf{-typical} iff $\abs{r(X)-H(p)} \le \epsilon$. 
\end{defn}
\begin{thm} 
Let $T_{N,\epsilon}$ be the set of $\epsilon$-typical sequences, then the following hold. 
\begin{enumerate}[(i)]
\item $\lim_{N \to \infty}P\set{X \in T_{N,\epsilon}} = 1$. 
\item For large enough $N$, we have $2^{N(H(p)-\epsilon)} \le \abs{T_{N,\epsilon}} \le 2^{N(H(p)+\epsilon)}$. 
\item For any $x \in T_{N,\epsilon}$, we have $2^{-N(H(p)+\epsilon)} \le P\set{X = x} \le 2^{-N(H(p)-\epsilon)}$.
\end{enumerate}
\end{thm}
\begin{proof} 
From Corollary to the Sanov's theorem, we have $P\set{X \notin T_{N,\epsilon}} \ee e^{-NI}$, 
where the exponent 
\EQ{
I = \min\set{D(q \Vert p): q \notin K}, \text{ where } K =\set{q \in \sM(\sX): \abs{\E_qr(X) -H(p)}  \le \epsilon}.
} 
\begin{enumerate}[(i)]
\item That is, we can write $K = \set{q \in \sM(\sX): \abs{D(q\Vert p)+ H(q)- H(p)} \le \epsilon}$. 
It follows that $p \in K$ and hence $I > 0$. 
Therefore, $\lim_{N \to \infty}P\set{X \notin T_{N,\epsilon}} = 0$. 
\item The compact set of types of the sequence $X \in T_{N,\epsilon}$ is given by $K$. 
Let $q \in K$, then we have $\abs{D(q\Vert p)+ H(q)- H(p)} \le \epsilon$, which implies that $\abs{H(q) - H(p)} \le \epsilon$. 
From the previous proposition, we have 
\EQ{
2^{N(H(p)-\epsilon)} \le \abs{T_{N,\epsilon}} \le 2^{N(H(p)+\epsilon)}.
}
\item Recall that $T_{N,\epsilon} = \set{x \in \sX^N: \abs{r(x)- H(p)} \le \epsilon}$, and hence 
for any $x \in T_{N,\epsilon}$ we have $P\set{X =x} = 2^{-Nr(x)}$. 
The result follows. 
\end{enumerate}
\end{proof}

\begin{defn} 
The behavior described in the above theorem is called the \textbf{asymptotic equipartition property}. 
\end{defn}

\section{Correlated variables}
For independent random variables in finite spaces, 
the probability of a large deviation is easily computed by combinatorics. 
We now present some general result for large deviations of non-independent random variables using  Legendre transforms and saddle point methods. 

\subsection{Legendre transformation} 
Consider the joint distribution of a set of $N$ random variables over the configuration space $\sX^N$ given by 
\EQ{
P_N(x) = P_N(x_1, \dots, x_N),~ x \in \sX^N. 
}
Let $f: \sX \to \R$ be a real valued function, and its empirical average 
\EQ{
\of(X) = \frac{1}{N}\sum_{i=1}^Nf(X_i). 
}
We showed that finite fluctuation of $\of$ is exponentially unlikely for \emph{i.i.d.} random variables. 
We will show the same for `weakly correlated' random variables. 
In particular, let $P_N(x)$ be the Boltzmann distribution of $N$ particle interacting system, and $f$ be a macroscopic observable. 
Then, we will show that the relative fluctuation of macroscopic observables is small. 
\begin{assum}
The distribution of $\of$ follows a \textbf{large-deviation principle}, meaning that the asymptotic behavior of the distribution at large $N$ is 
\EQ{
P_N(\of) \ee e^{-N I(\of)},
}
with a \textbf{rate function} $I(\of) \ge 0$. 
\end{assum}
In order to determine the rate function, a useful method is to `tilt' the measure $P_N(\cdot)$ in such
a way that the rare events responsible for $O(1)$ fluctuations of $\of$ become likely. 
\begin{defn} 
The \textbf{logarithmic moment-generating function} of $\of$ is defined as 
\EQ{
\psi_N(t) \triangleq \frac{1}{N}\log\expect{e^{Nt\of(x)}},~ t \in \R.
}
\end{defn}
When the large-deviation principle holds, we can evaluate the large-$N$ limit of $\psi_N(t)$ using the saddle point method
\EQ{
\psi(t) \triangleq \lim_{N \to \infty}\psi_N(t) = \lim_{N \to \infty}\frac{1}{N}\log\int d\of e^{-NI(\of)}e^{Nt\of}.
}
It follows that $\psi(t)$ is the Legendre transform of $I(\of)$
\EQ{
\psi(t) = \sup\set{t\of - I(\of): \of \in \R}. 
}
Since $\psi(t)$ is supremum of affine functions in $t$, it is convex in $t$. 
Therefore, we can invert the Legendre transform as 
\EQ{
I_\psi(\of) = \sup\set{t\of - \psi(t): t \in \R},
}
where $I_\psi(\of)$ is the convex envelope of $I(\of)$. 
This procedure is useful when computing $\psi(t)$ is easier than the probability distribution $P_N(\of)$.

\end{document}